SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-06-25 paper/paper.tex
final paper edits.

% can be 9pt
\documentclass{sigplanconf}
\usepackage{url}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{color}
\usepackage{url}
\usepackage{alltt}
\usepackage{tabularx}
\usepackage{subfigure}
\usepackage{fancyhdr}
\usepackage{paralist}
\usepackage{multirow}
\usepackage{pgf}
\usepackage{nth}

@@

\newcommand{\xmonad}{\textsc{xmonad}}

@@

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc,monadic}.  Originally designed for Haskell,
QuickCheck has been ported to other languages and is a now a widely-used testing
tool.  Because QuickCheck generates random values for testing, counterexamples
it finds may be substantially larger than a minimal counterexample.  In their
original QuickCheck paper~\cite{qc}, the authors report the following user
experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}

That is, automated testing helps find a counterexample, but it may be a very
large counterexample.  We are left facing the \emph{shrinking problem}: assume
there exist inputs $i_0, \ i_1, \ldots, \ i_n$, such that for program
$P$ (modeled as a function mapping inputs to output) and property $\phi$,
$\phi(P(i_0, \ i_1, \ldots, \ i_n))$ is false.  The shrinking problem is to find
the smallest inputs $i'_0, \ i'_1, \ldots, \ i'_n$, according to some
measure, such that $\phi(P(i'_0, \ i'_1, \ldots, \ i'_n))$ is also false.

In general, the problem is too expensive to solve optimally, since it requires
showing that for a set of failing inputs, there exist no strictly smaller inputs
that also cause the property to fail.  Such a demonstration requires exhaustive
testing or proof.  Solutions to the shrinking problem are therefore approximated
to efficiently discover sufficiently small inputs causing the input to fail.
Usually, the shrunk inputs are derived algorithmically (but perhaps
nondeterministically) from a known larger set of inputs resulting in failure,
and the inputs may not be minimal.

@@

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples.  This leap of
abstraction is often implicitly made by the programmer in determining the reason
why counterexamples fail the property.  For example, Figure~\ref{fig:xmonadex}
contains a relatively small counterexample returned when QuickChecking a
property in (a bug-injected version of) \xmonad, a popular X11 window manager
written in Haskell~\cite{xmonad}.  (This counterexample uses Haskell's default
\ttp{Show} instances, which uses record syntax.)  Programmers may be familiar
with having to debug a ``wall of text'' as shown above.  What if instead a
formula like the following were returned, stating that for any well-typed values
$x_0$, $x_1$, $x_2$, and $x_3$, tested, a counterexample is found?

@@

\paragraph{SmartCheck}
Motivated by the problems of reducing and generalizing large counterexamples, we
developed SmartCheck.  SmartCheck takes a counterexample produced by some
oracle and generically minimizes and generalizes the counterexample.

@@

After presenting some preliminary definitions in
Section~\ref{sec:preliminaries}, in Section~\ref{sec:shrinking}, we describe
SmartCheck's generic counterexample reduction algorithm.  The algorithm is
particularly novel in that it generates new random sub-values during the
reduction phase (see related work in Section~\ref{sec:related}).

@@

In this section, we motivate in more detail the challenges in shrinking
counterexamples by comparing manual approaches using QuickCheck to SmartCheck.
(We focus on shrinking rather than generalization here since counterexample
generalization is unique to SmartCheck.)  We will show how a small data type and
simple property can result in large counterexamples without any shrinking.  Then
we show the difficulty in designing an efficient shrink implementation manually,
which is required when using QuickCheck; we will show a poor
design before arriving at a ``canonical'' manual solution.%%   However, even the
%% canonical solution produces significantly larger counterexamples and takes
%% significantly longer to produce them than SmartCheck, even though SmartCheck's
%% shrinking strategy is completely automatic and generic.

@@

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T i0 i1 i2 i3 i4) =
  [ T a b c d e | a <- shrink i0, b <- shrink i1
                , c <- shrink i2, d <- shrink i3
                , e <- shrink i4 ]
\end{code}
%
\noindent
While the above definition appears reasonable on first glance, there are two
problems with it.  First, the result of (\ttp{shrink t}) is null if any list
contained in \ttp{t} is null, in which case \ttp{t} will not be shrunk.  More
troublesome is that with QuickCheck's default \ttp{Arbitrary} instances, the
length of potential counterexamples returned by \ttp{shrink} can easily exceed
$10^{10}$, which is an intractable number of tests for QuickCheck to analyze.
The reason for the blowup is that shrinking a list \ttp{[a]} produces a list
\ttp{[[a]]}, a list of lists; each list is a new shrinking of the original list.
Then, we take the cross-product of the generated lists.  For the example above,
with some counterexamples, the shrinking stage may appear to execute for hours,
consuming ever more memory, without returning a result.

@@

In early 2014, QuickCheck added the ability to derive \ttp{shrink} instances for
algebraic data type using the same generic package that SmartCheck does; see
Section~\ref{sec:preliminaries}.  (The first release of the SmartCheck software
was mid-2012.)  With this new feature, the user declares
%
\begin{code}
shrink = genericShrink
\end{code}
%
\noindent
to obtain efficient shrink instances.

@@

index tree 3  = (Just . SubVal) (B\sub{4} L\sub{6} L\sub{8})
index tree 12 = Nothing

@@

subVals (B\sub{0} (B\sub{1} L\sub{0} L\sub{1}) L\sub{2}) = SubVal
  (Node (B\sub{0} (B\sub{1} L\sub{0} L\sub{1}) L\sub{2})
        [ Node (B\sub{1} L\sub{0} L\sub{1}) [Node L\sub{0} [], Node L\sub{1} []]
        , Node L\sub{2} []
        ])
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 4th sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.  (Rather than indexes into a data-structure, an
alternative representation is to use a zipper data structure~\cite{zipper} to
traverse data.  We have chosen explicit indexes to write simple tail-recursive
algorithms that can easily be transcribed to imperative languages.)

@@

The hypothesis may be incorrect for some properties, in which case SmartCheck
may potentially fail to discover a smaller counterexample.  However, in
Sections~\ref{sec:existential} and~\ref{sec:precondition}, we describe
approaches to generalize counterexamples based on discovering new counterexample
variants.  These generalization techniques are executed in an (optional)
generalization phase, run after the reduction phase, in which this hypothesis is
implemented.

@@

For example, consider a simple calculator language containing constants,
addition, and division, together with an evaluator that checks if the divisor is
0 and returning \ttp{Nothing} in that case:

\pagebreak

%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Now consider the property \ttp{prop\_div}, claiming that if \ttp{divSubTerms}
holds on an expression, then the evaluator returns \ttp{Just} a value:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
divSubTerms (Div e0 e1)   =  divSubTerms e0
                          && divSubTerms e1

@@

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if no new
values satisfy the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is roughly the dual of \ttp{pass} in the reduction algorithm: (\ttp{fail
  prop cex}) returns (\ttp{Just True}) if \ttp{cex} passes \ttp{prop}'s
precondition but fails the property; (\ttp{Just False}) if \ttp{cex}
non-trivially satisfies \ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails
\ttp{prop}'s precondition.

Like in the reduction algorithm user-specified flags bound the behavior of the
algorithm.  We bound the size of values to generate by the flag \ttp{scMaxSize},
which is independent of the size of the particular sub-value.  The flag
\ttp{scMaxForall} is the analogue of the \ttp{scMaxReduce} flag, determining the
number of values generated in trying to generalize a value.  The flag
\ttp{scMinForall} is the minimum number of \ttp{Just False} results required
from \ttp{fail} to extrapolate from failed tests to a universal claim.  So, for
example, if \ttp{scMaxForall} is set to 30 and \ttp{scMinForall} is set to 20,
we generate 30 new values, 20 of which must pass the precondition but fail the
property to claim the counterexample can be generalized.

The algorithm's complexity is $\mathcal{O}(n)$, where $n$ is the number of
constructors in the counterexample.  Again, we assume that the cost for each
generating random values and testing them at each index is constant.

@@

Unlike universal sub-value generalization, existential sub-value generalization
is sound.  The existential claim is only that for each variant, there exists at
least one counterexample, so inductive reasoning is not required.

@@

To understand the shape matching algorithm, consider a few values of type
\ttp{Exp}, defined in Section~\ref{sec:subval}:
%
\begin{code}
e0 = Div (C 1) (C 2)
e1 = Div (C 1) (C 3)
e2 = Div (Add (C 1) (C 2)) (C 7)
e3 = Div (Div (C 8) (C 2)) (C 7)
\end{code}
%
\noindent
Then the following hold:
%
\begin{code}
matchesShape e0 (e1, [])  == True
matchesShape e1 (e2, [])  == False
matchesShape e1 (e2, [0]) == True
matchesShape e3 (e2, [0]) == True
\end{code}
%
The first equation holds because we ignore opaque types, so the arguments to
\ttp{C} are considered equal.  The second equation fails because at index 0,
\ttp{e1} contains the constructor \ttp{C} and \ttp{e2} contains the constructor
\ttp{Add}.  If, however, index 0 has been generalized, then we ignore the sub-value
at index 0, and they match (the third equation).  The same reasoning holds for
the fourth equation.

@@

SmartCheck generically operates over arbitrary algebraic data and so uses a
generics library to encode generic traversals.  Specifically, SmartCheck uses
``GHC Generics'', a recently-developed generics library for
Haskell~\cite{generics}.  The generics library allows algebraic data types to
automatically derive the type class \ttp{SubTypes} presented in
Section~\ref{sec:preliminaries}.  One limitation with GHC Generics is that it
does not (currently) support generalized algebraic data types~\cite{gadts}.

@@

\section{Experiments}\label{sec:experiments}
We describe two experiments using SmartCheck, including an \xmonad property and a
property about a natural language processing library.

@@

\xmonad passes all of its QuickCheck tests, but let us see what might happen to
a new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  An existing deletion function in
\xmonad exists, which is quite complex, given the amount of state that is
managed by \ttp{StackSet}.  However, one function used in deletion is to filter
the stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (/= w) \}
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (== w) \}
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
QuickCheck returns the large value shown in the introduction.  That value is a
relatively small counterexample, but even the smallest \ttp{StackSet} values are
somewhat visually overwhelming due to the number of fields within it.  Recall
the value returned by SmartCheck after generalization:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
Let us examine what was generalized.  In our test run, we chose to treat data
maps as opaque, so the fourth element of \ttp{StackSet} is not generalized, but
is simply the empty map, which looks uninteresting.  The second and third fields of
\ttp{StackSet} are generalized, but the first one is not.  There is something
particular about it.  So the culprit is one of the small constants (1 and -1) or
having a \ttp{Just} value rather than a \ttp{Nothing}: it turns out that what
matters is having a \ttp{Just} value, which is the stack field that deletion
works on!

@@

  \item \emph{Parser}, a parser/pretty-printer for a toy imperative language
    containing a parser bug that switches the arguments of an binary expression
    operator.
\end{itemize}

@@

\multirow{2}{*}{Parser}
  & QC &  4   & 0.024 & 0  & 0.020 & 4 & 0.061 \\\cline{2-8}
  & SC &  6   & 0.116 & 2  & 0.083 & 9 & 0.230 \\\hline

@@

The \emph{Reverse} benchmark essentially provides a lower-bound on the benefit
shrinking in general, since the original counterexamples are generally close to
being minimal.  Surprisingly, SmartCheck slightly outperforms QuickCheck in
efficiency.  The other three benchmarks have larger counterexamples, so the
benefit of shrinking is more pronounced.

SmartCheck finds smaller counterexamples in the \emph{Div0} and \emph{Heap}
benchmarks, while QuickCheck shrinking finds smaller counterexamples faster in
the \emph{Parser} example.  The \emph{Parser} example is one in which
SmartCheck's strategy of generating random values during counterexample
reduction is not a benefit, and so this is why it likely under performs a
generic \ttp{shrink} method.  Moreover, because SmartCheck uses QuickCheck as a
front-end, the two approaches are complementary: QuickCheck shrinking can be
used instead of or in combination with SmartCheck reduction.

@@

HDD is most similar to SmartCheck's reduction algorithm, with an important
difference: HDD (and DD) is deterministic, so the algorithm only succeeds in
reducing the counterexample only if a new counterexample can be constructed from
the original one.  Our approach combines the speed of delta debugging with the
power of QuickCheck to \emph{randomly} discover structurally smaller
counterexamples.  The idea of randomization in test-case reduction was developed
at approximately the same time as SmartCheck and first reported in the domain of
reducing C~programs that demonstrate compiler bugs~\cite{regehr}.  We believe our
work is the first to explore the technique in a more general setting.
Furthermore, we believe ours is the first to explore the idea of counterexample
generalization.

@@

Finally, SmartCheck bears some similarity to QuickSpec, a testing-based library
that infers equational properties about programs~\cite{qs} insofar as they both
attempt to generalize counterexamples based on specific inputs.  QuickSpec
attempts to infer equational properties of programs through random testing.
Similarly, Daikon infers assertions for C, C++, Java, and Perl by observing
relationships between variables in executions of a program~\cite{daikon}.
SmartCheck does not attempt to infer properties like these tools do, so it is in
one sense less general.  However, SmartCheck's three kinds of counterexample
generalization are novel.

>>>>

% can be 9pt
\documentclass{sigplanconf}
\usepackage{url}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{color}
\usepackage{url}
\usepackage{alltt}
\usepackage{tabularx}
\usepackage{subfigure}
\usepackage{fancyhdr}
\usepackage{paralist}
\usepackage{multirow}
\usepackage{pgf}
\usepackage{nth}
\usepackage{xspace}

@@

\newcommand{\xmonad}{\textsc{xmonad}\xspace}

@@

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
property-based testing~\cite{qc,monadic}.  Originally designed for Haskell,
QuickCheck has been ported to other languages and is a now a widely-used testing
tool.  Because QuickCheck generates random values for testing, counterexamples
it finds may be substantially larger than a minimal counterexample.  In their
original QuickCheck paper~\cite{qc}, the authors report the following user
experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}

That is, automated testing helps find a counterexample, but it may be a very
large counterexample.  We are left facing the \emph{shrinking problem}: consider
program $P$ modeled as a function mapping an input vector $\vec{i}$ to an output
value.  Consider property $\phi$ modeled as a function mapping the inputs of $P$
to a Boolean value that has the form $pre(\vec{i}) \Longrightarrow
post(P(\vec{i}))$, where $pre$ is a precondition on inputs and $post$ is a
postcondition on the program output.  Let $<$ be a partial order over the set of
possible inputs.  Given inputs $\vec{i}$ such that $\phi(\vec{i})$ is false, the
shrinking problem is to find inputs $\vec{j}$ such that $\phi(\vec{j})$ is false
and there exist no inputs $\vec{k}$ such that $\vec{k} < \vec{j}$ and
$\phi(\vec{k})$ is false.

In general, the problem is too expensive to solve optimally, since it requires
showing that for a set of failing inputs, there exist no strictly smaller inputs
that also cause the property to fail.  Such a demonstration requires exhaustive
testing or proof.  Solutions to the shrinking problem are therefore approximated
to efficiently discover sufficiently small inputs causing the input to fail.
Usually, the shrunk inputs are derived algorithmically (but perhaps
nondeterministically) from known larger inputs resulting in failure.

@@

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples characterizing the bug.
This leap of abstraction is often implicitly made by the programmer in
determining the reason why counterexamples fail the property.  For example,
Figure~\ref{fig:xmonadex} contains a relatively small counterexample returned
when using QuickCheck to test a property in (a bug-injected version of) \xmonad,
a popular X11 window manager written in Haskell~\cite{xmonad}.  (This
counterexample uses Haskell's default \ttp{Show} instances, which uses record
syntax.)  Programmers may be familiar with having to debug a ``wall of text'' as
shown above.  What if instead a formula like the following were returned,
stating that for any well-typed values $x_0$, $x_1$, $x_2$, and $x_3$, tested, a
counterexample is found?

@@

\paragraph{SmartCheck}
Motivated by the problems of reducing and generalizing large counterexamples, we
developed SmartCheck.  SmartCheck takes a counterexample produced by some oracle
and generically minimizes and generalizes the counterexample.  After presenting
some preliminary definitions in Section~\ref{sec:preliminaries}, in
Section~\ref{sec:shrinking}, we describe SmartCheck's generic counterexample
reduction algorithm.

@@

@@

In this section, we motivate in more detail the challenges in shrinking
counterexamples by comparing manual approaches using QuickCheck to SmartCheck.
(We focus on shrinking rather than generalization here since counterexample
generalization is unique to SmartCheck.)  We will show how a small data type and
simple property can result in large counterexamples without any shrinking.  Then
we show the difficulty in designing an efficient shrink implementation.  We will
show a poor design before arriving at a ``canonical'' manual solution.

%% %%
%% However, even the
%% canonical solution produces significantly larger counterexamples and takes
%% significantly longer to produce them than SmartCheck, even though SmartCheck's
%% shrinking strategy is completely automatic and generic.

@@

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T i0 i1 i2 i3 i4) =
  [ T a b c d e | a <- shrink i0, b <- shrink i1
                , c <- shrink i2, d <- shrink i3
                , e <- shrink i4 ]
\end{code}
%
\noindent
While the above definition appears reasonable on first glance, there are two
problems with it.  First, the result of (\ttp{shrink t}) is null if any list
contained in \ttp{t} is null, in which case \ttp{t} will not be shrunk.  More
troublesome is that with QuickCheck's default \ttp{Arbitrary} instances, the
length of potential counterexamples returned by \ttp{shrink} can exceed
$10^{10}$, which is an intractable number of tests for QuickCheck to analyze.
The reason for the blowup is that shrinking a list \ttp{[a]} produces a list
\ttp{[[a]]}, a list of lists; each list is a new shrinking of the original list.
Then, we take the cross-product of the generated lists.  For the example above,
with some counterexamples, the shrinking stage may appear to execute for hours,
consuming ever more memory, without returning a result.

@@

In early 2014, QuickCheck added the ability to derive \ttp{shrink} instances for
algebraic data types.  (The first release of the SmartCheck software was
mid-2012.)  With this new feature, the user declares
%
\begin{code}
shrink = genericShrink
\end{code}
%
\noindent
to obtain efficient shrink instances.

@@

index tree 0  = tree
index tree 3  = (Just . SubVal) (B\sub{4} L\sub{6} L\sub{8})
index tree 12 = Nothing

@@

subVals (B\sub{0} (B\sub{1} L\sub{0} L\sub{1}) L\sub{2}) = SubVal
  (Node (B\sub{0} (B\sub{1} L\sub{0} L\sub{1}) L\sub{2})
        [ Node (B\sub{1} L\sub{0} L\sub{1}) [Node L\sub{0} [], Node L\sub{1} []]
        , Node L\sub{2} []
        ])
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 3rd sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.  (Rather than indexes into a data-structure, an
alternative representation is to use a zipper data structure~\cite{zipper} to
traverse data.  We have chosen explicit indexes to write simple tail-recursive
algorithms that can easily be transcribed to imperative languages.)

@@

Incidentally, QuickCheck's generic shrink implementation is fundamentally build
around the variant counterexample hypothesis.  For a given counterexample
\ttp{cex}, smaller candidates produced by \ttp{shrink} contain only the
constructors found in \ttp{cex} as opposed to just the outermost constructor, as
in SmartCheck.  Sometimes shrinking gets ``stuck'' at a local minimum due to a
lack of entropy in generating smaller candidates.

The hypothesis may be unsuitable for some properties, in which case SmartCheck
(and especially QuickCheck) may potentially fail to discover a smaller
counterexample.  However, in Sections~\ref{sec:existential}
and~\ref{sec:precondition}, we describe approaches to generalize counterexamples
based on discovering new counterexample variants.  These generalization
techniques are executed in an (optional) generalization phase, run after the
reduction phase, in which this hypothesis is implemented.

@@

For example, consider a simple calculator language containing constants,
addition, and division, together with an evaluator that checks if the divisor is
0 and returning \ttp{Nothing} in that case:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Now consider the property \ttp{prop\_div}, claiming that if \ttp{divSubTerms}
holds on an expression, then the evaluator returns \ttp{Just} a value:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
divSubTerms (Div e0 e1)   =  divSubTerms e0
                          && divSubTerms e1

@@

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if no new
values satisfy the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is roughly the dual of \ttp{pass} in the reduction algorithm: (\ttp{fail
  prop cex}) returns (\ttp{Just True}) if \ttp{cex} passes \ttp{prop}'s
precondition but fails the property, (\ttp{Just False}) if \ttp{cex}
non-trivially satisfies \ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails
\ttp{prop}'s precondition.

Like in the reduction algorithm, user-specified flags bound the behavior of the
algorithm.  We bound the size of values to generate by the flag \ttp{scMaxSize},
which is independent of the size of the particular sub-value.  The flag
\ttp{scMaxForall} is the analogue of the \ttp{scMaxReduce} flag, determining the
number of values generated in trying to generalize a value.  The flag
\ttp{scMinForall} is the minimum number of \ttp{Just False} results required
from \ttp{fail} to extrapolate from failed tests to a universal claim.  So, for
example, if \ttp{scMaxForall} is set to 30 and \ttp{scMinForall} is set to 20,
we generate 30 new values, 20 of which must pass the precondition but fail the
property to claim the counterexample can be generalized.

The algorithm's complexity is $\mathcal{O}(n)$, where $n$ is the number of
constructors in the counterexample.  Again, we assume that the cost for
generating random values and testing them at each index is constant.

@@

Unlike universal sub-value generalization, existential sub-value generalization
is sound.  The existential claim is only that for each variant, there exists at
least one counterexample.

@@

To understand the shape matching algorithm, consider a few values of type
\ttp{Exp}, defined in Section~\ref{sec:subval}:
%
\begin{code}
e0 = Div (C 1) (C 2)
e1 = Div (C 1) (C 3)
e2 = Div (Add (C 1) (C 2)) (C 7)
e3 = Div (Div (C 8) (C 2)) (C 7)
\end{code}
%
\noindent
Then the following hold:
%
\begin{code}
matchesShape e0 (e1, [])  == True
matchesShape e1 (e2, [])  == False
matchesShape e1 (e2, [1]) == True
matchesShape e3 (e2, [1]) == True
\end{code}
%
The first equation holds because we ignore opaque types, so the arguments to
\ttp{C} are considered equal.  The second equation fails because at index 1,
\ttp{e1} contains the constructor \ttp{C} and \ttp{e2} contains the constructor
\ttp{Add}.  If, however, index 1 has been generalized, then we ignore the sub-value
at index 1, and they match (the third equation).  The same reasoning holds for
the fourth equation.

@@

SmartCheck generically operates over arbitrary algebraic data and so uses a
generics library to encode generic traversals.  Specifically, SmartCheck uses
``GHC Generics'', a generics library for Haskell~\cite{generics}.  The library
allows algebraic data types to automatically derive the type class
\ttp{SubTypes} presented in Section~\ref{sec:preliminaries}.  One limitation
with GHC Generics is that it does not (currently) support generalized algebraic
data types~\cite{gadts}.

@@

\section{Experiments}\label{sec:experiments}
We describe two experiments using SmartCheck, including an \xmonad property and a
property about a natural language processing library.  Then we present a small
set of benchmarks comparing SmartCheck and QuickCheck.

@@

\xmonad passes all of its QuickCheck tests, but let us see what might happen to
a new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  An existing deletion function in
\xmonad exists, which is quite complex, given the amount of state that is
managed by \ttp{StackSet}.  However, one function used in deletion is to filter
the stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (/= w) \}
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (== w) \}
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
QuickCheck returns the large value shown in Figure~\ref{fig:xmonadex}.  That
value is a relatively small counterexample, but even the smallest \ttp{StackSet}
values are somewhat visually overwhelming due to the number of fields within it.
Recall the value returned by SmartCheck after generalization:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
Let us examine what was generalized.  In our test run, we chose to treat data
maps as opaque, so the fourth element of \ttp{StackSet} is not generalized, but
is simply the empty map, which looks uninteresting.  The second and third fields of
\ttp{StackSet} are generalized, but the first one is not.  There is something
particular about it.  So the culprit is one of the small constants (1 and -1) or
having a \ttp{Just} value rather than a \ttp{Nothing}: it turns out that what
matters is having a \ttp{Just} value, which is the stack field that deletion
works on!

@@

  \item \emph{Parser}, a parser/pretty-printer for a toy imperative language
    containing a parser bug that switches the arguments of disjunction
    expressions.
\end{itemize}

@@

\multirow{2}{*}{Parser}
  & QC &  4   & 0.010 & 0  & 0.006 & 4  & 0.023 \\\cline{2-8}
  & SC &  7   & 0.182 & 3  & 0.124 & 12 & 0.418 \\\hline

@@

The \emph{Reverse} benchmark essentially provides a lower-bound on the benefit of
shrinking in general, since the original counterexamples are generally close to
being minimal.  Surprisingly, SmartCheck slightly outperforms QuickCheck in
efficiency.  The other three benchmarks have larger counterexamples, so the
benefit of shrinking is more pronounced.

SmartCheck finds smaller counterexamples in the \emph{Div0} and \emph{Heap}
benchmarks, while QuickCheck shrinking finds smaller counterexamples faster in
the \emph{Parser} example.  The example is one in which SmartCheck's
counterexample reduction strategy is less optimal than QuickCheck's.  Recall
from Section~\ref{sec:variant} that QuickCheck's generic shrink implementation
generates candidates that contain a subset of the constructors from original
counterexample.  In the parser example, the bug is localized in the
counterexample, arising from a single expression in the program.  SmartCheck
wastes effort generating new programs using new constructors. SmartCheck is
better suited, however, at avoiding local minima for other properties and
programs.

@@

HDD is most similar to SmartCheck's reduction algorithm, with an important
difference: HDD (and DD) is deterministic, so the algorithm only succeeds in
reducing the counterexample only if a new counterexample can be constructed from
the original one.  Our approach combines the speed of delta debugging with the
power of QuickCheck to \emph{randomly} discover structurally smaller
counterexamples.  The idea of randomization in test-case reduction was
independently developed at approximately the same time as SmartCheck and first
published in the domain of reducing C~programs that demonstrate compiler
bugs~\cite{regehr}. We believe our work is the first to explore
the idea of counterexample generalization.

@@

Finally, SmartCheck bears some similarity to QuickSpec, a testing-based library
that infers equational properties about programs~\cite{qs} insofar as they both
attempt to generalize counterexamples based on specific inputs.  QuickSpec
attempts to infer equational properties of programs through random testing.
Similarly, Daikon infers assertions for C, C++, Java, and Perl by observing
relationships between variables in executions of a program~\cite{daikon}.
SmartCheck does not attempt to infer properties like these tools dox.
